Indexing Variation Graphs\thanks{This work was supported by the Wellcome Trust grant 098051.}

### 2016-08-14 paper/paper.tex
Almost final submission.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{textgreek}
\usepackage{microtype}
\usepackage{hyperref}

@@

There are also several structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015,Na2016}) using graphs as a space-efficient way of indexing similar sequences. While the problem is different from indexing the paths in a graph, the techniques used are similar.

@@

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a \kmer{k} index of a variation graph. The de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is encoded with a generalization of the FM\nobreakdash-index. GCSA2 often uses around 1~bit per \kmer{k} with large values of $k$. Explicit de~Bruijn graphs cannot achieve similar space usage, as they store some information for each \kmer{k}. GCSA2 also includes extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph. The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

@@

\subsection{Text indexes}

The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting substring queries. It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges. If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$. Although fast and versatile, suffix trees are impractical with large texts, as they require much more space than the text itself.

@@

\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{itemize}
\item each node $v_{X} \in V$ represents a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
\end{itemize}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

@@

\subsection{FM-index for graphs}

@@

\subsection{De Bruijn graphs as path indexes}

@@

\subsection{Path graphs}

@@

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$. Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$. The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where
\begin{itemize}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i) \in V$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{itemize}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

@@

\subsection{Pruned de Bruijn graphs}

@@

\begin{figure}[t!]
\includegraphics[width=\textwidth]{gcsa2_pruned_index.pdf}
\caption{Left: An \protect\orderk{3} pruned de~Bruijn graph $G''$ \protect\kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}. Right: GCSA for graph $G''$. Leftward arrows show how backward searching proceeds, with the red arrows demonstrating it from pattern $\mathtt{T}$ to pattern $\mathtt{AT}$. Rightward arrows tell where the samples for each node are located, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
\end{figure}

@@

As in the original GCSA, we sort the nodes of the path graph in lexicographic order, encode the indegrees and outdegrees in bitvectors $\bvIN$ and $\bvOUT$, and store the predecessor labels in $\BWT$. See Figure~\ref{figure:pruned-index} for an example. If lexicographic range $[sp_{i+1}, ep_{i+1}]$ matches suffix $X[i+1, \abs{X}-1]$ of pattern $X$, we can find the range matching suffix $X[i, \abs{X}-1]$ as
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
In order to support $\locate$ queries, we \emph{sample} the values of a node, if the node has multiple incoming edges or the values cannot be derived from the values of the predecessor. We may also sample other values to improve query performance. The sampled nodes are marked in bitvector $B_{S}$, the number of values in each sample is encoded in unary in bitvector $B_{V}$, and the sampled values are stored in integer array $V_{S}$. A detailed description of the data structure can be found in Appendix~\ref{appendix:gcsa2}.

@@

GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\href{https://github.com/jltsiren/gcsa2}{https://github.com/jltsiren/gcsa2}). As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. The index uses fast succinct data structures instead of the smaller but slower compressed data structures for most of its components. The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0. All files were stored on a distributed Lustre file system. We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.

The vg \emph{variation graphs} have strings as node labels. Each node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples.

@@

Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes. We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache. The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them. Index size grows even more slowly with $k$, as we sample more values in lower order graphs. For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph. Extensions from Appendix~\ref{appendix:extensions} increase index size to 1.12~bits per \kmer{k}.

\begin{table}[t!]
\begin{center}
\caption{GCSA2 construction with different orders of the path graph. Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
\begin{tabular}{c|ccc|cc}
\hline
\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
\hline
 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}[t!]
\begin{center}
\caption{Index sizes with different orders $k$ of the path graph. Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
\begin{tabular}{c|cc|c|c|c}
\hline
\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
\hline
 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
128 &  116G & 5.73G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
\hline
\end{tabular}
\end{center}
\end{table}

We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread. We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference. The results can be seen in Table~\ref{table:benchmark}. The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components. A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

@@

\section{Conclusions}

GCSA2 is a path index for variation graphs. It uses a de~Bruijn graph as an index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index. The index supports queries of length up to $k$ exactly and longer queries with false positives. GCSA2 includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}). The index has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.

We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions. Queries are several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.

The primary desing goals in GCSA2 have been query performance and index size. The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece, and provides several options for dealing with complex regions. Other path indexes work with a more restricted class of graphs \cite{Siren2014,Huang2013,Kim2015-2016,Maciuca2016}, are at least an order of magnitude slower \cite{Huang2013,Maciuca2016}, require much more space \cite{Bowe2012,Pell2012,Cazaux2014,Marcus2014}, or are theoretical proposals that have never been implemented \cite{Thachuk2013}.

The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice. As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently. We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.

\paragraph*{Acknowledgements.}

The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.

\bibliographystyle{plainurl}
\bibliography{paper}

@@

For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
\end{proof}

@@

Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ is a prefix of key $G'.\gkey(u')$. If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$. Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
\end{proof}

@@

(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$.
\end{proof}

\begin{proof}[Lemma~\ref{lemma:dbg-fp}: No short false positives]
We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.

Let $v' \in G'.\find(X)$ be a node. By Lemma~\ref{lemma:pg-keys}, pattern $X$ prefix-matches key $G'.\gkey(v')$. For every node $v \in G'.\gvalue(v')$, there is a substring $S[i, i+k-1] = G'.\gkey(v'_{0})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$, with $\mathcal{S}.\gnode(S, i) = v$. Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$.
\end{proof}

@@

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$. Therefore substring $X[j] \cdot G''.\gkey(v''_{j+1})$ exists in $\mathcal{S}$.

@@

Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$. We can also start from path $P''$ and determine the node $v_{0}$. As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
\end{proof}

@@

The above is true for all nodes $x' \in G'.\gpred(v', G'.\glabel(u'))$. Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
\end{proof}

@@

Checking whether key $G''.\gkey(u'')$ prefix-matches string $c \cdot G''.\gkey(v'')$ can be done by using the GCSA for de~Bruijn graph $G_{d} = (V, E)$ and the last character array $L$. If the lexicographic rank of \kmer{k} $X$ is $i$, the lexicographic rank of \kmer{k} $(cX)[0, k-1]$ is $G_{d}.\LF(i, c)$. If \kmer{2k} $X$ is encoded with \kmer{k} ranks $(i, j)$, we can encode string $cX$ as a lexicographic range of \kmer{k} rank sequences, with $(G_{d}.\LF(i, c), G_{d}.\LF(j, L[i]), G_{d}.\LF(0, L[j]))$ as the lower bound and $G_{d}.\LF(\abs{V_{d}}-1, L[j])$ as the last rank in the upper bound. String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only the corresponding lexicographic ranges of \kmer{k} rank sequences overlap.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{textgreek}
\usepackage{microtype}

% hyperref does not work with LNCS
%\usepackage{hyperref}
\usepackage{url}

@@

There are also structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015,Na2016}) using graphs as a space-efficient way of indexing similar sequences. While the problem is different from indexing the paths in a graph, the techniques used are similar.

@@

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a \kmer{k} index of a variation graph. The de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is encoded with a generalization of the FM\nobreakdash-index. GCSA2 often uses less space (e.g.~1~bit per \kmer{k}) than other de~Bruijn graph-based indexes, which have to store some information for each \kmer{k} explicitly. The index also includes extensions based on compressed suffix trees. The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

@@

\subsection{Text Indexes}

The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting substring queries. It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges. Although fast and versatile, suffix trees are impractical with large texts, as they require much more space than the text itself.

@@

\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{itemize}
\item each node $v_{X} \in V$ represents a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
\end{itemize}
We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
\end{definition}

@@

\subsection{FM-Index for Graphs}

@@

\subsection{De Bruijn Graphs as Path Indexes}

@@

\subsection{Path Graphs}

@@

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$. Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$. The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where
\begin{itemize}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i) \in V$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{itemize}
We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
\end{definition}

@@

\subsection{Pruned de Bruijn Graphs}

@@

\begin{figure}[t!]
\includegraphics[width=\textwidth]{gcsa2_pruned_index.pdf}
\caption{Left: An \protect\orderk{3} pruned de~Bruijn graph $G''$ \protect\kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}. Right: GCSA for graph $G''$. Leftward arrows show how backward searching works, with the red arrows showing it from pattern $\mathtt{T}$ to pattern $\mathtt{AT}$. Rightward arrows match nodes with sampled values, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
\end{figure}

@@

As in the original GCSA, we sort the nodes of the path graph in lexicographic order, encode the indegrees and outdegrees in bitvectors $\bvIN$ and $\bvOUT$, and store the predecessor labels in $\BWT$. See Figure~\ref{figure:pruned-index} for an example. If lexicographic range $[sp_{i+1}, ep_{i+1}]$ matches suffix $X[i+1, \abs{X}-1]$ of pattern $X$, we can find the range matching suffix $X[i, \abs{X}-1]$ as
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
In order to support $\locate$ queries, we \emph{sample} the values of a node, if the node has multiple incoming edges or the values cannot be derived from the predecessor. We may also sample other values to improve query performance. The sampled nodes are marked in bitvector $B_{S}$, the number of values in each sample is encoded in unary in bitvector $B_{V}$, and the sampled values are stored in integer array $V_{S}$. A detailed description of the data structure can be found in Appendix~\ref{appendix:gcsa2}.

@@

GCSA2 is the path indexing library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). Most index components are encoded using fast non-compressed bitvectors. As GCSA2 uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.

We used a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory for the experiments, and stored all files on a distributed Lustre file system. The system was running Ubuntu~12.04 on Linux kernel~3.2.0. We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.

\subsection{Construction}

The vg \emph{variation graphs} have strings as node labels. A node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations. For indexing, the graph is implicitly converted into an input graph with single-character labels. We always sample the input graph nodes corresponding to the initial offsets of variation graph node labels.

@@

Tables~\ref{table:construction} and \ref{table:indexes} show construction requirements and index sizes, respectively. We can build a whole-genome index overnight using less than 96~gigabytes of memory, including disk cache. The index contains $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the path graph only uses 4.4\nobreakdash--5.8~billion nodes to represent them. For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph. Extensions based on suffix trees increase the size to 1.12~bits per \kmer{k}.

\begin{table}[t!]
\begin{center}
\caption{GCSA2 construction. Order of the path graph; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
\begin{tabular}{c|ccc|cc}
\hline
\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
\hline
 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}[t!]
\begin{center}
\caption{GCSA2 index sizes. Order $k$ of the path graph; number of \kmer{k}s and nodes in the path graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
\begin{tabular}{c|cc|c|c|c}
\hline
\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
\hline
 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
128 &  116G & 5.73G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
\hline
\end{tabular}
\end{center}
\end{table}

Index construction uses more memory with $k = 32$ than with larger values of $k$. The \orderk{32} path graph has more nodes, where we cannot derive the values from the predecessor node. As we sample more values, we need more memory in the final phase of construction. With larger values of $k$, the path graph resembles the input graph better, and we sample less values. For the same reason, index size decreases with large values of $k$, even though the graph requires more space.

\subsection{Queries}

We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread. We also ran the same queries in an FM\nobreakdash-index of the reference. The results can be seen in Table~\ref{table:benchmark}. The FM\nobreakdash-index (\texttt{csa\_wt<>} with SA sample density 17) uses the same SDSL components as GCSA2, allowing us to compare the fundamental techniques and ignore most implementation-specific details.

@@

With \kmer{16}s, where both indexes match a similar number of patterns, $\find$ queries in GCSA2 are about 3x slower than in the FM\nobreakdash-index. With longer patterns, the difference grows larger, as the query stops early when the index cannot match the pattern. $\select$ queries on bitvector $\bvIN$ are the main source of the difference, as they are much slower than $\rank$ queries. A simplified GCSA encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

When the number of distinct occurrences is high, $\locate$ queries are about 5x slower than in the FM\nobreakdash-index. With less occurrences, unpredictable factors such as SA order sampling in the FM\nobreakdash-index or nodes with multiple values and filtering out duplicates in GCSA2 affect the results. Queries $\parent$ and $\countq$ are both clearly faster than a single step of backward searching.

\section{Discussion}

GCSA2 is a path index for variation graphs. It uses a de~Bruijn graph as a \kmer{k} index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index. The index supports queries of length up to $k$ exactly and longer queries with false positives. GCSA2 also includes extensions based on suffix trees; other extensions have been considered but not implemented (see Appendix~\ref{appendix:hypertext}). The index is used in the variation graph toolkit vg for e.g.~read alignment based on maximal exact matches.

We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions. Queries are several times slower than in a similar implementation of the FM\nobreakdash-index for sequences.

The primary desing goals in GCSA2 were query performance and index size. The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece, and provides several options for dealing with complex regions. Other path indexes work with a more restricted class of graphs \cite{Siren2014,Huang2013,Kim2015-2016,Maciuca2016}, are at least an order of magnitude slower \cite{Huang2013,Maciuca2016}, require much more space \cite{Bowe2012,Pell2012,Cazaux2014,Marcus2014}, or are theoretical proposals that have never been implemented \cite{Thachuk2013}.

We may want to determine whether a pattern matches known \emph{haplotypes} or only their recombinations. As GCSA2 does not support this directly, vg checks it afterwards using a separate structure \cite{Novak2016}. The \emph{FM\nobreakdash-index of alignment} \cite{Na2015,Na2016} embeds the haplotypes directly in a GCSA-like index and reports the haplotypes matching the $\find$ query. While the solution depends on the specific properties of the graph, it could be possible to extend it to work with any GCSA.

% Maybe in the full version
%The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice. As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently. We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.

\paragraph{Acknowledgements.} The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.

\bibliographystyle{splncs03}
\bibliography{paper}

@@

For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$. \qed
\end{proof}

@@

Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ is a prefix of key $G'.\gkey(u')$. If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$. Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$. \qed
\end{proof}

@@

(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$. \qed
\end{proof}

\begin{proof}[Lemma~\ref{lemma:dbg-fp}: No short false positives]
We may assume without loss of generality that graph $G'$ is a de~Bruijn graph. Let $v' \in G'.\find(X)$ be a node. By Lemma~\ref{lemma:pg-keys}, pattern $X$ prefix-matches key $G'.\gkey(v')$. For every node $v \in G'.\gvalue(v')$, there is a substring $S[i, i+k-1] = G'.\gkey(v'_{0})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$, with $\mathcal{S}.\gnode(S, i) = v$. Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$. \qed
\end{proof}

@@

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$. Therefore substring $X[j] \cdot G''.\gkey(v''_{j+1})$ exists in $\mathcal{S}$.

@@

Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$. We can also start from path $P''$ and determine the node $v_{0}$. As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$. \qed
\end{proof}

@@

The above is true for all nodes $x' \in G'.\gpred(v', G'.\glabel(u'))$. Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}. \qed
\end{proof}

@@

Checking whether key $G''.\gkey(u'')$ prefix-matches string $c \cdot G''.\gkey(v'')$ can be done by using the GCSA for de~Bruijn graph $G_{d} = (V_{d}, E_{d})$ and the last character array $L$. If the lexicographic rank of \kmer{k} $X$ is $i$, the lexicographic rank of \kmer{k} $(cX)[0, k-1]$ is $G_{d}.\LF(i, c)$. If \kmer{2k} $X$ is encoded with \kmer{k} ranks $(i, j)$, we can encode string $cX$ as a lexicographic range of \kmer{k} rank sequences, with $(G_{d}.\LF(i, c), G_{d}.\LF(j, L[i]), G_{d}.\LF(0, L[j]))$ as the lower bound and $G_{d}.\LF(\abs{V_{d}}-1, L[j])$ as the last rank in the upper bound. String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only the corresponding lexicographic ranges of \kmer{k} rank sequences overlap.
